---
title: "Part-2"
author: 'Florencia Luque  and Seyed Amirhossein Mosaddad '
output: pdf_file
date: "2024-10-18"
---
## Part 2


```{r,warning=FALSE}
library(caret)
library(caretEnsemble)
library(h2o)
library(tidymodels)
library(ROCR)
```

### Relation between variables

#### Stroke with Gender

```{r,warning=FALSE}
tab_st_gd = table(stroke_data$stroke,stroke_data$gender)
chisq.test(tab_st_gd)
```
The *p-value* is larger than 0.05 this mean that there's not evidence of dependency between the variables gender and stroke. 

#### Stroke with Hypertension

```{r,warning=FALSE}
tab_st_hy = table(stroke_data$stroke,stroke_data$hypertension)
chisq.test(tab_st_hy)
```
The *p-value* is a lot smaller than 0.05. This mean that there's a relation between getting a stroke and hypertension. This is a could be a comprobation of the hypothesis that we established earlier about the existence of a relation between this two variables.

#### Stroke with Heart Disease

```{r,warning=FALSE}
tab_st_hd = table(stroke_data$stroke,stroke_data$heart_disease)
chisq.test(tab_st_hd)
```
The *p-value* is a lot smaller than 0.05. This mean that there's a relation between getting a stroke and heart disease This is a could be a comprobation of the hypothesis that we established earlier about the existence of a relation between this two variables.


#### Stroke with Residence Type

```{r,warning=FALSE}
tab_st_rt = table(stroke_data$stroke,stroke_data$Residence_type)
chisq.test(tab_st_rt)
```
As we had seen in the graph there´s no evidence to say that there's a relation between the type of residence and getting a stroke.

#### Stroke with Ever Married

```{r,warning=FALSE}
tab_st_em = table(stroke_data$stroke,stroke_data$ever_married)
chisq.test(tab_st_em)
```
Apparently there's a relation within this two variables. Having a stroke have a relation with have been or had been ever married. 

#### Stroke with Smoking Status

```{r,warning=FALSE}
tab_st_sk = table(stroke_data$stroke,stroke_data$smoking_status)
chisq.test(tab_st_sk)
```
The *p-value* is a lot smaller than 0.05 so there's is a relation between the variables. This was something that we *dont know what to write here*


#### Stroke with Work Type

```{r,warning=FALSE}
tab_st_wt = table(stroke_data$stroke,stroke_data$work_type)
chisq.test(tab_st_wt)
```
There's a relation between the variables (*p-value*<0.05). This we think was because of the difference between the quantity of people who got a stroke and work independently and the people who work with children because the difference was big between them.

#### Heart Disease and Hypertension

```{r,warning=FALSE}
tab_hy_hd = table(stroke_data$heart_disease,stroke_data$hypertension)
chisq.test(tab_hy_hd)
```
There's a relation between hypertension and heart disease and both variable are related to stroke. This could be a good indicator that within only one of this variables we could have the same information in the model.

#### Heart Disease and Ever Married

```{r,warning=FALSE}
tab_hd_em = table(stroke_data$heart_disease,stroke_data$ever_married)
chisq.test(tab_hd_em)
```
#### Heart Disease and Smoking Status

```{r,warning=FALSE}
tab_hd_sk= table(stroke_data$heart_disease,stroke_data$smoking_status)
chisq.test(tab_hd_sk)
```

#### Heart Disease and Work Type

```{r,warning=FALSE}
tab_hd_wt = table(stroke_data$heart_disease,stroke_data$work_type)
chisq.test(tab_hd_wt)
```
#### Hypertension and Ever Married

```{r,warning=FALSE}
tab_hy_em = table(stroke_data$heart_disease,stroke_data$ever_married)
chisq.test(tab_hy_em)
```

#### Hypertension and Smoking Status

```{r,warning=FALSE}
tab_hy_sk = table(stroke_data$heart_disease,stroke_data$smoking_status)
chisq.test(tab_hy_sk)
```

#### Hypertension and Work Type

```{r,warning=FALSE}
tab_hy_wt= table(stroke_data$heart_disease,stroke_data$work_type)
chisq.test(tab_hy_wt)
```

# Perform a t-test
t.test(continuous_var ~ dichotomous_var, data = your_data)
for continuous with categorical 

1) Logistic Regression #flo

2) VSM #soroush

3) Xgboost for classification 0-1 #soroush

For classification using package sparsediscrim with tuning parameters:

Model (model, character)

```{r,warning=FALSE}
model_data = subset(stroke_data,select = -c(id,cat_weight,glucose_category))
log_reg_model = glm(stroke~age+hypertension+heart_disease,data = model_data,family = "binomial")
summary(log_reg_model)
```


Split the data to get a 80% for training and 20% for testing.#flo

```{r,warning =FALSE}
head(stroke_data)
h2o.init()
stroke_h2o=as.h2o(stroke_data)
split_data = h2o.splitFrame(data=stroke_h2o,ratios=0.8,seed=23)
train = split_data[[1]]
test = split_data[[2]]
pred = c("gender","age","hypertension","heart_disease","ever_married","work_type","Residence_type","avg_glucose_level","bmi","smoking_status")
aml = h2o.automl(x=pred,y="stroke",training_frame=train,max_models=30,seed=23)
lb <- aml@leaderboard
print(lb, n = nrow(lb))

```