---
title: "Stroke dataset"
author: "Florencia Luque  and Seyed Amirhossein Mosaddad "
date: "October 6, 2024"
output: html_document
---

```{r,echo=FALSE}
library(ggplot2)
library(readr)
library(e1071)
library(dplyr)
library(formattable)
library(pander)
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)

```

## Introduction

This dataset is a data obtain from *kaggle* and is used to predict if a pacient will probably get a stroke based on characteristic of them like gender, age, bmi, glucose levels.\

The stroke variable have a 4.8% of people have had one. We want to check the distributions of the variables and possibles explanations of which variable can make an impact to get a stroke before creating a model to proved or been proved wrong about it.\

The data have the follow variables:\

1) id: unique identifier
2) gender: "Male", "Female" or "Other"
3) age: age of the patient
4) hypertension: 0 if the patient doesn't have hypertension, 1 if the patient has hypertension
5) heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the patient has a heart disease
6) ever_married: "No" or "Yes"
7) work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed"
8) Residence_type: "Rural" or "Urban"
9) avg_glucose_level: average glucose level in blood
10) bmi: body mass index
11) smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
12) stroke: 1 if the patient had a stroke or 0 if not

The next is a summary of the data.

```{r setup, include=FALSE}
path = "C:\\Users\\flore\\Desktop\\UC3M\\programacion en R\\proyect\\stroke data\\"
data = read.csv("healthcare-dataset-stroke-data.csv",header = TRUE)
str(data)
data$gender = as.factor(data$gender)
data$hypertension = as.factor(data$hypertension)
data$heart_disease = as.factor(data$heart_disease)
data$work_type = as.factor(data$work_type)
data$Residence_type = as.factor(data$Residence_type)
data$ever_married = as.factor(data$ever_married)
data$smoking_status = as.factor(data$smoking_status)
data$stroke = as.factor(data$stroke)
data$bmi =as.numeric(data$bmi)
stroke_counts <- table(data$stroke)
percentage_of_1 <- (stroke_counts["1"] / sum(stroke_counts)) * 100
```

```{r}
summary(data)
```


### Dataset
We will star the analysis with the categorical variables.

## Categorical Variables

### Gender
This variable has 2 categories Male and Female (there's one person who is Other but it's only one so we can't make any assumptions about this data).

The next table shows a summary of the quantity of people getting a stroke by gender and the corresponding percentage.

```{r, warning=FALSE}
stroke_gender = data %>% group_by(stroke,gender) %>% summarise(n = n(),.groups = "drop") %>% group_by(gender) %>% mutate(percent = n/sum(n)*100)

pander(stroke_gender)
```

```{r}
ggplot(stroke_gender, aes(x = "", y = percent, fill = factor(stroke))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~ gender) +  # Separate charts for each gender
  labs(title = "Stroke Distribution by Gender", fill = "Stroke", y = "Percentage") +
  scale_fill_manual(values = c("0" = "#249", "1" = "deeppink1")) +  # Custom colors for 0 and 1
   theme_void()
```

As you can see there's a higher % of male that have had a stroke in the data. This number is a little bit higher than the population so we don't think this would have a great impact in the future model.


### Hypertension

```{r, warning=FALSE}
pander(data %>% group_by(stroke,hypertension) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(stroke) %>% mutate(percent = n/sum(n)*100))

```

### Heart disease

```{r, warning=FALSE}
pander(data %>% group_by(stroke,heart_disease) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(heart_disease) %>% mutate(percent = n/sum(n)*100))

```

### Ever Married

```{r}
pander(data %>% group_by(stroke,ever_married) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(ever_married) %>% mutate(percent = n/sum(n)*100))

```

### Work Type

```{r}
pander(data %>% group_by(stroke,work_type) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(work_type) %>% mutate(percent = n/sum(n)*100))

```

### Resident Type

```{r}
pander(data %>% group_by(stroke,Residence_type) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(Residence_type) %>% mutate(percent = n/sum(n)*100))

```

This could mean something

### Smoking Status

```{r}
pander(data %>% group_by(stroke,smoking_status) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(smoking_status) %>% mutate(percent = n/sum(n)*100))

```

It's very clear that this variable should be important in the model \## Numeric Variables

### BMI

The first variable to evaluate will be the *bmi*. This variable is a metric that represent the relation between height and weight of a person. As you can see we have `r (201/5110)*100` % percentage of NA. However this is less than 5% so to treat this variable we will deleted all the rows with the NA in *bmi*

```{r}
summary(data$bmi)
```

Next, we will check the frecquency of the data using the cut between (0-18.5] as underweight, (18.5-24.9] as normal, (24.9-29.9] as overweight, (29.9-34.9] as obese and over this as extremely obese. This value leaves over 50% of our population in normal and overweight and almost 40% in the largest side in obese and extremely obese.

```{r}
data = na.omit(data)
frec_table <- cut(data$bmi, breaks = c(0, 18.5, 24.9, 29.9, 34.9, Inf), 
                labels = c("Underweight", "Normal", "Overweight", "Obese", "Extremely Obese"))
data$cat_weight = frec_table

# Create frequency table
bmi_freq_table <- table(frec_table)
print(bmi_freq_table)
bmi_rel_freq <- prop.table(bmi_freq_table)*100
print(bmi_rel_freq)


```

If we check this with a graph we can see that the graph looks a bit symetrical with an inclination to the right. This could it mean that the distribution is not a normal like could it seem. We are going to check the kurtosis and skewness shape to check if there´s a problem.

```{r, warning = FALSE}
ggplot(data, aes(x=bmi)) + 
  geom_histogram(color="black", fill="cadetblue1")+
  labs(title = "BMI Distribution")



bmi_mean <- mean(data$bmi)
bmi_median <- median(data$bmi)

# Density plot with a vertical line at the mean
ggplot(data, aes(x = data$bmi)) +
  geom_density(fill = "aquamarine1", alpha = 0.6) +
  geom_vline(aes(xintercept = bmi_mean), color = "dodgerblue2", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = bmi_median), color = "hotpink1", linetype = "dotted", size = 1) +
  labs(title = "Density Plot of BMI with Mean and Median", x = "BMI", y = "Density") +
  annotate("text", x = bmi_mean + 0.5, y = 0.02, label = paste("Mean =", round(bmi_mean, 2)), color = "dodgerblue2") +
  annotate("text", x = bmi_median - 0.5, y = 0.015, label = paste("Median =", round(bmi_median, 2)), color = "hotpink1") +
  theme_minimal()
```

```{r}
skewness(data$bmi)
kurtosis(data$bmi)
```

The values of the shape tell us that this is not a normal distribution. There's a lot ot people in the center of the data but the quantity of people with a extremely high BMI that change the weigh of the tail in the distribution.

Looking only at the graph we expected that the Kurtosis was close to 3 but a 3.35 shows that this data have a heavier tail than a normal distribution. This means that we have outliers in the data. In this case there's a lot of values over 30. If we combined the Skewness value of 1.05 and the Kurtosis this tell us that the the distribution in skewed to the right and that the heavier tails is to this side explaining the outliers.

To see if the outliers are correctly in the heavier side we can check the boxplot of the *bmi*

```{r}
ggplot(data, aes(y = data$bmi)) + 
  geom_boxplot(fill = "lightblue", color = "darkblue") + # Change box and border color
  labs(title = "Boxplot of BMI", y = "BMI Values") + # Custom title and labels
  theme_minimal()
```

As you can see with the boxplot we can make sure that the *bmi* data have outliers within the largest values. The problem with this data is that we can not be sure if this is a mistake in the part of measure or maybe exist people with those values.

#### BMI with gender

```{r}


data_sub_bmi_gender= data %>% group_by(gender,stroke) %>% summarise(mean = mean(bmi))
print(data_sub_bmi_gender)

ggplot(data, aes(x = data$bmi, fill = data$gender)) + 
  geom_histogram(position = "identity", bins = 30)
```

#### BMI with hypertension

```{r}


data_sub_bmi_hyper= data %>% group_by(hypertension,stroke) %>% summarise(mean = mean(bmi))
print(data_sub_bmi_hyper)
ggplot(data, aes(x = bmi, fill = hypertension)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

#### BMI with stroke

```{r, warning=FALSE}


data_sub_bmi_catw= data %>% group_by(cat_weight,stroke) %>% summarise(mean = mean(bmi))
print(data_sub_bmi_catw)

ggplot(data, aes(x = bmi, fill = stroke)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

### Glucose levels

#### Glucose levels with gender

```{r, warning=FALSE}
ggplot(data, aes(x = avg_glucose_level, fill = gender)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

#### Glucose levels with hypertension

```{r, warning=FALSE}


data_sub_glu_hyp= data %>% group_by(hypertension,stroke) %>% summarise(mean = mean(avg_glucose_level))
print(data_sub_glu_hyp)
ggplot(data, aes(x = avg_glucose_level, fill = hypertension)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

#### Glucose levels with stroke

```{r, warning=FALSE}
data_sub_glu= data %>% group_by(stroke) %>% summarise(mean = mean(avg_glucose_level))
print(data_sub_glu)

ggplot(data, aes(x = avg_glucose_level, fill = stroke)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

### Age

#### Age levels with gender

```{r, warning=FALSE}
data_sub_age_gender= data %>% group_by(gender,stroke) %>% summarise(mean = mean(age))
print(data_sub_age_gender)
ggplot(data, aes(x = age, fill = gender)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

#### Age levels with hypertension

```{r, warning=FALSE}

data_sub_age_hyp= data %>% group_by(hypertension,stroke) %>% summarise(mean = mean(age))
print(data_sub_age_hyp)
ggplot(data, aes(x = age, fill = hypertension)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```

jhjh \#### Age levels with stroke

```{r, warning=FALSE}

data_sub_age= data %>% group_by(stroke) %>% summarise(mean = mean(age))
print(data_sub_age)
ggplot(data, aes(x = age, fill = stroke)) + 
  geom_histogram(alpha = 0.6, position = "identity", bins = 30)
```
