---
title: "Stroke dataset"
author: "Florencia Luque  and Seyed Amirhossein Mosaddad "
date: "October 6, 2024"
output: pdf_document
---

```{r,echo=FALSE,warning=TRUE}
library(ggplot2)
library(readr)
library(e1071)
library(dplyr)
library(formattable)
library(pander)
library(agricolae)
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)

```

## Introduction

This dataset is a data obtain from *kaggle* and is used to predict if a pacient will probably get a stroke based on characteristic of them like gender, age, bmi, glucose levels.\

The stroke variable have a 4.8% of people have had one. We want to check the distributions of the variables and possibles explanations of which variable can make an impact to get a stroke before creating a model to proved or been proved wrong about it.\

The data have the follow variables:\

1) id: unique identifier
2) gender: "Male", "Female" or "Other"
3) age: age of the patient
4) hypertension: 0 if the patient doesn't have hypertension, 1 if the patient has hypertension
5) heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the patient has a heart disease
6) ever_married: "No" or "Yes"
7) work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed"
8) Residence_type: "Rural" or "Urban"
9) avg_glucose_level: average glucose level in blood
10) bmi: body mass index
11) smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
12) stroke: 1 if the patient had a stroke or 0 if not

The next is a summary of the data.

```{r setup, include=FALSE}
path = "C:\\Users\\flore\\Desktop\\UC3M\\programacion en R\\proyect\\stroke data\\"
data = read.csv("healthcare-dataset-stroke-data.csv",header = TRUE)
str(data)
data = data[data$gender!="Other",]
data$gender = as.factor(data$gender)
data$hypertension = as.factor(data$hypertension)
data$heart_disease = as.factor(data$heart_disease)
data$work_type = as.factor(data$work_type)
data$Residence_type = as.factor(data$Residence_type)
data$ever_married = as.factor(data$ever_married)
data$smoking_status = as.factor(data$smoking_status)
data$stroke = as.factor(data$stroke)
data$bmi =as.numeric(data$bmi)
stroke_counts <- table(data$stroke)
percentage_of_1 <- (stroke_counts["1"] / sum(stroke_counts)) * 100


```

```{r}
summary(data)
```


### Dataset
We will star the analysis with the categorical variables.

## Categorical Variables

### Gender
This variable has 2 categories Male and Female (there's one person who is Other but it's only one so we can't make any assumptions about this data).

The next table shows a summary of the quantity of people getting a stroke by gender and the corresponding percentage.

```{r, warning=FALSE}
stroke_gender = data %>% group_by(stroke,gender) %>% summarise(n = n(),.groups = "drop") %>% group_by(gender) %>% mutate(percent = n/sum(n)*100)

pander(stroke_gender)
```
```{r} 
ggplot(data)+aes(x=stroke,fill=gender) + geom_bar(position=position_dodge())+theme_minimal()

```
As you can see there's a higher % of male that have had a stroke in the data. This number is a little bit higher than the population so we don't think this would have a great impact in the future model.


### Hypertension

```{r, warning=FALSE}
pander(data %>% group_by(stroke,hypertension) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(hypertension) %>% mutate(percent = n/sum(n)*100))

```

```{r} 
ggplot(data)+aes(x=stroke,fill=hypertension) + geom_bar(position=position_dodge())+theme_minimal()

```
The % of people who have had a stroke and have hypertension are significantly highier than the population. There's a difference approx. 9%. This could mean that if you have hypertension you could be more likely to get a stroke. 
### Heart disease

```{r, warning=FALSE}
pander(data %>% group_by(stroke,heart_disease) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(heart_disease) %>% mutate(percent = n/sum(n)*100))

```

```{r} 
ggplot(data)+aes(x=stroke,fill=heart_disease) + geom_bar(position=position_dodge())+theme_minimal()

```
The data of the people who have a heart diseased look even more likely to get a stroke than the ones who have hypertension. Maybe there's a relation between hypertension and having a heart diseased. Heart diseased is a variable to check for more details.

### Ever Married

```{r}
pander(data %>% group_by(stroke,ever_married) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(ever_married) %>% mutate(percent = n/sum(n)*100))

```

```{r} 
ggplot(data)+aes(x=stroke,fill=ever_married) + geom_bar(position=position_dodge())+theme_minimal()

```
there's more people that have had a stroke you have been married than not been married. The difference with the population is big but with the ones that never married maybe not get married lower your changes or getting a stroke.
### Work Type

```{r}
pander(data %>% group_by(stroke,work_type) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(work_type) %>% mutate(percent = n/sum(n)*100))
```


```{r} 
ggplot(data)+aes(x=stroke,fill=work_type) + geom_bar(position=position_dodge())+theme_minimal()

```
Clairly the self employed have the highest % of people who have han a stroke. This could be because of stress and the people who work with children have the lowest. Hope that working with children reduces your chances.
### Resident Type

```{r}
pander(data %>% group_by(stroke,Residence_type) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(Residence_type) %>% mutate(percent = n/sum(n)*100))

```

```{r} 
ggplot(data)+aes(x=stroke,fill=Residence_type) + geom_bar(position=position_dodge())+theme_minimal()

```
The part of the city that you live looks like it doesn't matter because the % of stroke is very similar to the population and almost the same between the types of resident.
This could mean something

### Smoking Status

```{r}
pander(data %>% group_by(stroke,smoking_status) %>% summarise(n = n(),.groups = "drop")%>%
  group_by(smoking_status) %>% mutate(percent = n/sum(n)*100))

```

```{r} 
ggplot(data)+aes(x=stroke,fill=smoking_status) + geom_bar(position=position_dodge())+theme_minimal()

```
This results call the attention because the people you formerly smoked have 3% higher than the population but obe who smokes or never have similar numbers. This could be a multivariate problem because this type of data probably mix more than one variable at a time.
It's very clear that this variable should be important in the model \## Numeric Variables

### BMI

The first variable to evaluate will be the *bmi*. This variable is a metric that represent the relation between height and weight of a person. As you can see we have `r (201/5110)*100` % percentage of NA. However this is less than 5% so to treat this variable we will deleted all the rows with the NA in *bmi*

```{r}
summary(data$bmi)
```

Next, we will check the frecquency of the data using the cut between (0-18.5] as underweight, (18.5-24.9] as normal, (24.9-29.9] as overweight, (29.9-34.9] as obese and over this as extremely obese. This value leaves over 50% of our population in normal and overweight and almost 40% in the largest side in obese and extremely obese.

```{r}
data = na.omit(data)
frec_table <- cut(data$bmi, breaks = c(0, 18.5, 24.9, 29.9, 34.9, Inf), 
                labels = c("Underweight", "Normal", "Overweight", "Obese", "Extremely Obese"))
data$cat_weight = frec_table

# Create frequency table
bmi_freq_table <- table(frec_table)
print(bmi_freq_table)
bmi_rel_freq <- prop.table(bmi_freq_table)*100
print(bmi_rel_freq)


```

If we check this with a graph we can see that the graph looks a bit symetrical with an inclination to the right. This could it mean that the distribution is not a normal like could it seem. We are going to check the kurtosis and skewness shape to check if there´s a problem.
```{r,include=FALSE}
aux_bmi = hist(data$bmi)
```


```{r}
stat.freq(aux_bmi)
```


```{r, warning = FALSE}


ggplot(data, aes(x=bmi)) + 
  geom_histogram(color="black", fill="cadetblue1")+
  labs(title = "BMI Distribution")




bmi_mean <- mean(data$bmi)
bmi_median <- median(data$bmi)

# Density plot with a vertical line at the mean
ggplot(data, aes(x = data$bmi)) +
  geom_density(fill = "aquamarine1", alpha = 0.6) +
  geom_vline(aes(xintercept = bmi_mean), color = "dodgerblue2", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = bmi_median), color = "hotpink1", linetype = "dotted", size = 1) +
  labs(title = "Density Plot of BMI with Mean and Median", x = "BMI", y = "Density") +
  annotate("text", x = bmi_mean + 0.5, y = 0.02, label = paste("Mean =", round(bmi_mean, 2)), color = "dodgerblue2") +
  annotate("text", x = bmi_median - 0.5, y = 0.015, label = paste("Median =", round(bmi_median, 2)), color = "hotpink1") +
  theme_minimal()
```

```{r}
skewness(data$bmi)
kurtosis(data$bmi)
```

The values of the shape tell us that this is not a normal distribution. There's a lot ot people in the center of the data but the quantity of people with a extremely high BMI that change the weigh of the tail in the distribution.

Looking only at the graph we expected that the Kurtosis was close to 3 but a 3.35 shows that this data have a heavier tail than a normal distribution. This means that we have outliers in the data. In this case there's a lot of values over 30. If we combined the Skewness value of 1.05 and the Kurtosis this tell us that the the distribution in skewed to the right and that the heavier tails is to this side explaining the outliers.

To see if the outliers are correctly in the heavier side we can check the boxplot of the *bmi*

```{r}
ggplot(data, aes(y = data$bmi)) + 
  geom_boxplot(fill = "lightblue", color = "darkblue") + # Change box and border color
  labs(title = "Boxplot of BMI", y = "BMI Values") + # Custom title and labels
  theme_minimal()
```

As you can see with the boxplot we can make sure that the *bmi* data have outliers within the largest values. The problem with this data is that we can not be sure if this is a mistake in the part of measure or maybe exist people with those values. YOu can see this in the qqplot next.

```{r}
ggplot(data, aes(sample=bmi)) + stat_qq(size=2,color=rainbow(4908))+stat_qq_line()+theme_minimal()+
  labs(title = "QQ plot for bmi")

```

#### BMI with stroke

```{r, warning=FALSE}


data_sub_bmi_catw= data %>% group_by(cat_weight,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_catw)

ggplot(data, aes(x = bmi, fill = stroke)) + 
  geom_histogram(binwidth = 3)+
  facet_wrap(~stroke)+theme_minimal()+labs(title = "bmi histogram by stroke",x="bmi",y="count")

ggplot(data, aes(sample = bmi, fill = stroke))+
  stat_qq(size=2)+ stat_qq_line() +
  facet_wrap(~stroke)+theme_minimal()+labs(title = "bmi qqplot by stroke",x="stroke",y="count")

ggplot(data, aes(y = bmi, fill =stroke))+
  geom_boxplot()+
  facet_wrap(~gender)+theme_minimal()+labs(title = "bmi boxplot by stroke",x="stroke",y="count")
```


#### BMI with gender

```{r,warning=FALSE}
#gender

data_sub_bmi_gender= data %>% group_by(gender,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_gender)


ggplot(data, aes(y = bmi, fill = gender))+
  geom_boxplot()+
  facet_wrap(~gender)+theme_minimal()+labs(title = "bmi boxplot by gender",x="gender",y="count")


```

#### BMI with hypertension

```{r,warning=FALSE}
#hypertension

data_sub_bmi_hyper= data %>% group_by(hypertension,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_hyper)

ggplot(data, aes(y = bmi, fill =hypertension))+
  geom_boxplot()+
  facet_wrap(~gender)+theme_minimal()+labs(title = "bmi boxplot by hypertension",x="hypertension",y="count")
```

#### BMI by heart diseased

```{r, warning=FALSE}


data_sub_bmi_hd= data %>% group_by(heart_disease,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_hd)


ggplot(data, aes(y = bmi, fill =heart_disease))+
  geom_boxplot()+
  facet_wrap(~gender)+theme_minimal()+labs(title = "bmi boxplot by heart disease",x="heart disease",y="count")
```

#### BMI by ever married

```{r, warning=FALSE}


data_sub_bmi_em= data %>% group_by(ever_married,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_em)


ggplot(data, aes(y = bmi, fill =ever_married))+
  geom_boxplot()+
  facet_wrap(~ever_married)+theme_minimal()+labs(title = "bmi boxplot by ever married",x="ever married",y="count")
```


#### BMI by work type


```{r, warning=FALSE}


data_sub_bmi_wt= data %>% group_by(work_type,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_wt)


ggplot(data, aes(y = bmi, fill =stroke))+
  geom_boxplot()+
  facet_wrap(~work_type)+theme_minimal()+labs(title = "bmi boxplot by stroke",x="work type",y="count")
```

#### BMI by resident type

```{r, warning=FALSE}


data_sub_bmi_tr= data %>% group_by(Residence_type,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_tr)

ggplot(data, aes(y = bmi, fill =Residence_type))+
  geom_boxplot()+
  facet_wrap(~Residence_type)+theme_minimal()+labs(title = "bmi boxplot by resident type",x="resident type",y="count")
```


#### BMI by smoking status

```{r, warning=FALSE}


data_sub_bmi_catw= data %>% group_by(smoking_status,stroke) %>% summarise(mean = mean(bmi),.groups = "drop")
print(data_sub_bmi_catw)


ggplot(data, aes(y = bmi, fill =smoking_status))+
  geom_boxplot()+
  facet_wrap(~smoking_status)+theme_minimal()+labs(title = "bmi boxplot by smoking status",x="smoking status",y="count")
```

### Glucose levels

```{r}



glc_mean <- mean(data$avg_glucose_level)
glc_median <- median(data$avg_glucose_level)

# Density plot with a vertical line at the mean
ggplot(data, aes(x = avg_glucose_level)) +
  geom_density(fill = "aquamarine1", alpha = 0.6) +
  geom_vline(aes(xintercept = glc_mean), color = "dodgerblue2", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = glc_median), color = "hotpink1", linetype = "dotted", size = 1) +
  labs(title = "Density Plot of avg glucose with Mean and Median", x = "Avg. Glucose", y = "Density") +
  annotate("text", x = glc_mean + 0.5, y = 0.02, label = paste("Mean =", round(glc_mean, 2)), color = "dodgerblue2") +
  annotate("text", x =glc_median - 0.5, y = 0.015, label = paste("Median =", round(glc_median, 2)), color = "hotpink1") +
  theme_minimal()
```

ffd
```{r}
skewness(data$avg_glucose_level)
kurtosis(data$avg_glucose_level)
```

djfj

```{r}
ggplot(data, aes(y = avg_glucose_level)) + 
  geom_boxplot(fill = "lightblue", color = "darkblue") + # Change box and border color
  labs(title = "Boxplot of avg. glucose", y = "avg. glucose Values") + # Custom title and labels
  theme_minimal()
```

fdskjs

```{r}
ggplot(data, aes(sample=avg_glucose_level)) + stat_qq(size=2,color=rainbow(4908))+stat_qq_line()+theme_minimal()+
  labs(title = "QQ plot for avg. glucose")

```


#### Avg. Glucose levels with stroke

```{r, warning=FALSE}


data_sub_glu_st= data %>% group_by(stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_st)

ggplot(data, aes(x = avg_glucose_level, fill = stroke)) + 
  geom_histogram(binwidth = 3)+
  facet_wrap(~stroke)+theme_minimal()+labs(title = "avg. glucose histogram by stroke",x="avg. glucose",y="count")

ggplot(data, aes(sample = avg_glucose_level, fill = stroke))+
  stat_qq(size=2)+ stat_qq_line() +
  facet_wrap(~stroke)+theme_minimal()+labs(title = "avg. glucose qqplot by stroke",x="stroke",y="count")

ggplot(data, aes(y = avg_glucose_level, fill =stroke))+
  geom_boxplot()+
  facet_wrap(~gender)+theme_minimal()+labs(title = "avg. glucose boxplot by stroke",x="stroke",y="count")
```


#### Avg. Glucose levels with gender

```{r, warning=FALSE}
data_sub_glu_gd= data %>% group_by(gender,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_gd)

ggplot(data, aes(x = avg_glucose_level, fill = gender)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~gender)+theme_minimal()+labs(title = "avg glucose histogram by gender",x="avg glucose",y="count")

```

#### Glucose levels with hypertension

```{r, warning=FALSE}


data_sub_glu_hyp= data %>% group_by(hypertension,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_hyp)
ggplot(data, aes(x = avg_glucose_level, fill = hypertension)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~hypertension)+theme_minimal()+labs(title = "avg glucose histogram by hypertension",x="avg glucose",y="count")
```

#### Avg. Glucose levels with heart disease

```{r}
data_sub_glu_hd= data %>% group_by(heart_disease,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_hd)
ggplot(data, aes(x = avg_glucose_level, fill = heart_disease)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~heart_disease)+theme_minimal()+labs(title = "avg glucose histogram by heart disease",x="avg glucose",y="count")
```

#### Avg. Glucose levels with ever married
```{r}
data_sub_glu_em= data %>% group_by(ever_married,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_em)
ggplot(data, aes(x = avg_glucose_level, fill = ever_married)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~ever_married)+theme_minimal()+labs(title = "avg glucose histogram by ever married",x="avg glucose",y="count")

``` 

#### Avg. Glucose levels with work type

```{r}
data_sub_glu_wt= data %>% group_by(work_type,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_wt)
ggplot(data, aes(x = avg_glucose_level, fill = work_type)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~work_type)+theme_minimal()+labs(title = "avg glucose histogram by work type",x="avg glucose",y="count")
```

#### Avg. Glucose levels with resident type

```{r}
data_sub_glu_rt= data %>% group_by(Residence_type,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_rt)
ggplot(data, aes(x = avg_glucose_level, fill = Residence_type)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~Residence_type)+theme_minimal()+labs(title = "avg glucose histogram by resident type",x="avg glucose",y="count")
```


#### Avg. Glucose levels with smoking status

```{r}
data_sub_glu_st= data %>% group_by(smoking_status,stroke) %>% summarise(mean = mean(avg_glucose_level),.groups = "drop")
print(data_sub_glu_st)
ggplot(data, aes(x = avg_glucose_level, fill = smoking_status)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~smoking_status)+theme_minimal()+labs(title = "avg glucose histogram by smoking status",x="avg glucose",y="count")
```

### Age

```{r, warning = FALSE}


ggplot(data, aes(x=age)) + 
  geom_histogram(color="black", fill="cadetblue1")+
  labs(title = "Age Distribution")


age_mean <- mean(data$age)
age_median <- median(data$age)

# Density plot with a vertical line at the mean
ggplot(data, aes(x = data$age)) +
  geom_density(fill = "aquamarine1", alpha = 0.6) +
  geom_vline(aes(xintercept = age_mean), color = "dodgerblue2", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = age_median), color = "hotpink1", linetype = "dotted", size = 1) +
  labs(title = "Density Plot of BMI with Mean and Median", x = "BMI", y = "Density") +
  annotate("text", x = age_mean + 0.5, y = 0.02, label = paste("Mean =", round(age_mean, 0)), color = "dodgerblue2") +
  annotate("text", x = age_median - 0.5, y = 0.015, label = paste("Median =", round(age_median, 0)), color = "hotpink1") +
  theme_minimal()
```
#### Age levels with stroke

```{r, warning=FALSE}

data_sub_age= data %>% group_by(stroke) %>% summarise(mean = mean(age),.groups = "drop")
print(data_sub_age)
ggplot(data, aes(x = age, fill = stroke)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~stroke)+theme_minimal()+labs(title = "age histogram by stroke",x="age",y="count")
```




#### Age levels with gender

```{r, warning=FALSE}
data_sub_age_gender= data %>% group_by(gender,stroke) %>% summarise(mean = mean(age),.groups = "drop")
print(data_sub_age_gender)
ggplot(data, aes(x = age, fill = gender)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~gender)+theme_minimal()+labs(title = "age histogram by gender",x="age",y="count")
```

#### Age levels with hypertension

```{r, warning=FALSE}

data_sub_age_hyp= data %>% group_by(hypertension,stroke) %>% summarise(mean = mean(age),.groups = "drop")
print(data_sub_age_hyp)
ggplot(data, aes(x = age, fill = hypertension)) + 
  geom_histogram(binwidth = 5.5)+
  facet_wrap(~hypertension)+theme_minimal()+labs(title = "age histogram by hypertension",x="age",y="count")
```


